### 2020-06-01 HarvardX_Data_Science_Capstone/capstone.Rmd
Update files

---
title:
  Concrete compressive strength prediction with machine learning
date: "April 22, 2020"
author: "Pedro Bernardino Alves Moreira"
indent: true
abstract:
 Compressive strength is the main characteristic of concrete. This work used a set of samples widely used to build 6 forecast models of this characteristic, each one specifically for an age, 3, 7, 14, 28, 56 and 100 days from 7 ingredients (water, cement, fine aggregate, coarse aggregate, fly ash, blast furnace slag, and superplasticizers) and 6 relationships between the main ingredients (water/cement, coarse aggregate/cement, fine aggregate/cement, fine aggregate / coarse aggregate, water / coarse aggregate and water/fine aggregate) to make the predictions. The algorithm *Parallel Random Forest* was used (*parRF* of the *caret* package in *R*) . The results were satisfactory, being the best for the 3-day model (*RMSE* of 3.31) and the worst for the 100-day model (*RMSE* of 5.85). The 28-day model had *RMSE* of 4.72. The work indicated that the separation in a model for each age is promising and was able to generate compatible results or better than other studies, opening way for new models creation following this proposal, but using different algorithms or the combination of different algorithms.
output:
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc_depth: 2
bibliography: references.bib
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{supertabular}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage[table]{xcolor}
- \usepackage{wrapfig}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
- \usepackage{caption}
- \usepackage{floatrow}
- \floatsetup[table]{capposition=top}
- \floatsetup[figure]{capposition=top}
- \captionsetup{options=chunk}
- \DeclareNewFloatType{chunk}{placement=H, fileext=chk, name=}
- \renewcommand{\thechunk}{\arabic{chunk}}
- \usepackage{indentfirst}
- \usepackage{sectsty} \sectionfont{\centering}
---
```{r setup, include=FALSE, cache=F}
# Install libraries, if not already installed
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(caret)) install.packages("caret")
if(!require(knitr)) install.packages("knitr")
if(!require(kableExtra)) install.packages("kableExtra")
if(!require(gdata)) install.packages("gdata")
if(!require(dplyr)) install.packages("dplyr")
if(!require(tidyr)) install.packages("tidyr")
if(!require(cowplot)) install.packages("cowplot")
if(!require(pastecs)) install.packages("pastecs")
if(!require(factoextra)) install.packages("factoextra")
if(!require(reshape2)) install.packages("reshape2")
if(!require(purrr)) install.packages("purrr")
if(!require(gridExtra)) install.packages("gridExtra")
if(!require(questionr)) install.packages("questionr")
# Load libraries
library(tidyverse)
library(caret)
library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(gdata)
library(reshape2)
library(cowplot)
library(pastecs)
library(factoextra)
library(purrr)
library(gridExtra)
library(questionr)
# Set global options
options(pillar.sigfig = 5)
options(knitr.kable.NA = '')
opts_knit$set(eval.after = "fig.cap")
opts_chunk$set(echo = TRUE, fig.align = 'center', cache=T, prompt=F, highlight=T )
oldSource <- knit_hooks$get("source")
knit_hooks$set(source = function(x, options) {
  x <- oldSource(x, options)
  x <- ifelse(!is.null(options$ref), paste0("\\label{", options$ref,"}", x), x)
  ifelse(!is.null(options$codecap), paste0("\\captionof{chunk}{", options$codecap,"}", x), x)
})
```
\newpage
\tableofcontents
\newpage
\listoffigures
\listoftables
\newpage

@@

## Obtaining the data
The data was downloaded from the University of California Irvine website [@downloadData] (\ref{show-download-data}). In total there are 1030 samples with 9 columns. The samples were renamed and an id column was added to facilitate data manipulation (\ref{show-rename-dat-cols}). The columns were reordered to put the new id column in the first position (\ref{show-reorder-dat}). The first samples can be viewed in the table \ref{tab: first-samples}.

@@

```{r download-data, echo=F, include=F}
# Data download
url_base <- "https://archive.ics.uci.edu"
url <-  "/ml/machine-learning-databases/concrete/compressive/Concrete_Data.xls"
download.file(paste0(url_base, url), "data.xls")
dat <- read.xls("data.xls")
n_inicial_samples <- nrow(dat)
colnames(dat)
```
```{r rename-dat-cols, echo=F}
# Ranaming the columns
colnames(dat) <- c(
  "cement",
  "blast_furnace_slag",
  "fly_ash",
  "water",
  "superplasticizers",
  "coarse_aggregate",
  "fine_aggregate",
  "day",
  "mpa"
  )
dat$id <- seq.int(nrow(dat))
```
```{r reorder-dat, echo=F}
# Reordering the data
col_order <- c(
  "id",
  "cement",
  "blast_furnace_slag",
  "fly_ash",
  "water",
  "superplasticizers",
  "coarse_aggregate",
  "fine_aggregate",
  "day",
  "mpa"
)
dat <- dat[, col_order]
```
```{r col-names-and-units, echo=F}
# Defining column names and units
colNames <- c("ID", "Cement", "B.F.S.", "Fly ash", "Water",
             "Superp.", "C.Aggregate", "F.Aggregate", "Day", "Comp.Str.")
dfUnits <- c("", "$kg/m^3$", "$kg/m^3$","$kg/m^3$","$kg/m^3$",
             "$kg/m^3$","$kg/m^3$","$kg/m^3$","","$MPa$")
```
```{r first-samples, echo=F}
# Table First samples
caption <- "First 6 samples"
kable(
    dat[1:6,],
    col.names = dfUnits,
    escape = F,
    booktabs = T,
    caption = caption,
    linesep = "\\addlinespace",
    align = "c"
    ) %>%
    add_header_above(header = colNames, line = F, align = "c") %>%
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

@@

```{r removing-duplicated-samples, echo=F}
# Removing duplicated samples
n_distinct_samples <- dat %>% select(-c(id)) %>% n_distinct()
n_duplicated_samples <- n_inicial_samples - n_distinct_samples
dat <- dat[!duplicated(select(dat, -c(id))),]
n_samples <- nrow(dat)
```
The data show the variables in the columns and samples in the rows. However it was found that some samples are identical in proportions of ingredients, changing only the value of age and compressive strngth, for example, samples 653, 654, 678 and 681, shown in the table \ref{tab:similar-samples}.

@@

```{r similar-samples, echo=FALSE}
# Tabela - Samples with same composition
same_samples <- dat %>%
  filter(id %in% c(653, 678, 654, 681))
caption <- "Samples with same composition"
kable(
  same_samples[order(same_samples$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```
In addition, there are also samples with the same values and proportions of ingredients, but with different compressive strength, probably due to differences in the building process. This is the case, for example, of samples 472, 473 and 474, shown in the table  \ref{tab:similar-samples-2}.

@@

```{r similar-samples-2, echo=FALSE}
# Tabela - Same samples with different results
same_samples_2 <- dat %>%
  filter(id %in% c(472, 473, 474))
caption <- "Same samples with different results"
kable(
  same_samples_2[order(same_samples_2$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("scale_down"))
```
To facilitate the analysis of the samples, all samples that are the same in relation to the ingredients, have been assigned the same *id*. In addition, as the compressive strength at 28 days is the parameter to determine the grade of the concrete, only the elements containing that day among its samples were maintained. In the case of the same samples but with different results of compressive strength, the values were averaged. After all these changes (\ref{show-initial-data-cleaning}), the new total of samples was reduced to 970, containing 416 different settings for the proportions of ingredients.

@@

```{r similar-samples-same-id, echo=FALSE, include=TRUE}
# Tabela - Previous samples after processing
same_samples <- dat %>%
  filter(id == 653 | id == 472 & day == 28)
caption <- "Previous samples after processing"
kable(
  same_samples[order(same_samples$id, same_samples$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE,
  digits = 2
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

### Age selection
As previously described, the main age for analysis of compressive strength is 28 days, but other ages can also be used to build forecasting models. However, it is necessary to verify how relevant the data of these other ages are. Starting with the distribution of the samples in relation to each age  (\ref{show-boxplot}) shown in the figure \ref{fig:boxplot}.

@@

This hypothesis was verified using the principal component analysis method, applied to samples of these 3 ages (\ref{show-pca-90-91-100}). The figure \ref{fig:pca-90-91-100} shows how the samples relate to each other (which are similar or different) and revealed how each variable contributes to the analysis. The first two dimensions represent $ 37 \% $ and $ 24 \% $ respectively of the variance.

@@

```{r pca-90-91-100, echo=F, include=T, fig.cap = cap, fig.height=3.5}
# Figure - Principal component analysis - 90, 91 e 100 days
dat_90_91_100 <- dat %>%
  ungroup() %>%
  filter(day %in% c(90, 91, 100)) %>%
  select(-c(id, mpa))
cap <- "Principal component analysis - 90, 91 e 100 days"
colnames(dat_90_91_100) <- c(
   "Cem.", "B.F.S.", "Fly.A.","Water","Sup.","C.Ag.","F.Ag.","day")
pca <- prcomp(select(dat_90_91_100, -c(day)), scale = TRUE)
habillage <- dat_90_91_100$day
fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage=habillage,
  addEllipses = TRUE,
  ellipse.level=0.75) +
  ggtitle("") +
  theme_bw() +
  coord_cartesian(xlim = c(-3, 3.5), ylim = c(3, -5))
```
Another important point considered, of the concrete nature itself, is the fact that the growth rate of its compressive strength decreases with time, reaching a certain stability value. The figure \ref{fig:mpa-on-time} shows the compressive strength over the days for samples with more than 5 data, that is, data available for at least 6 different ages (\ref{show-mpa-on-team}).

@@

```{r mpa-on-time, echo=F, include=T, fig.cap=cap, fig.height=3}
# Figure - Compressive strength through time
cap <- "Compressive strength through time"
ylabel <- "MPa"
xlabel <- "Age (days)"
dat_duplicated_only <- dat %>%
  group_by(id) %>%
  filter(n()>5) %>%
  select(id, mpa, day)
dat_duplicated_only %>%
  ggplot(aes(day, mpa, fill=id, alpha = 0.5)) +
  geom_line() +
  xlab(xlabel) +
  ylab(ylabel) +
  theme_bw() +
  theme(legend.position = "none")
```
For the reasons presented in the figures \ref{fig:boxplot}, \ref{fig:pca-90-91-100} and \ref{fig:mpa-on-time}, it was considered that the ages of 90, 91 and 100 days can be grouped to improve reading and decrease sample noise. They were converted to the same value, the age of 100 days was chosen (\ref{show-join-90-91-100}. As shown in the figure \ref{fig:mpa-on-time}, the resistance only increases, after 100 days the resistance to compression will be greater than or equal to the value of 90 or 91 days.

@@

```{r join-90-91-100, echo=F}
# Joining 90, 91 and 100 days data
ind_90 <- dat$id[which(dat$day == "90")]
ind_91 <- dat$id[which(dat$day == "91")]
ind_100 <- dat$id[which(dat$day == "100")]
sum(duplicated(c(ind_90, ind_91, ind_100))) # 0
dat <- dat %>%
  ungroup() %>%
  mutate(day = ifelse(day %in% c(91, 90), 100, day))
```
Another topic analyzed in the selection of ages was the observed frequency of each age value after this transformation from 90, 91 in 100 days, shown in the figure \ref{fig:freq-ages}. Some values of days have very low concentrations of samples, at the risk of damaging more than helping to create the models, so they were removed (\ref{show-remove-ages-lower-50}). The criterion adopted was to maintain only ages with a frequency greater than 50, only the values of 3, 7, 14, 28, 56 and 100 days.

@@

### Data reorganization
The samples were grouped to maintain only a separate sample from each set of configuration of the proportions of the ingredients, adding new variables/columns for resistance at each age (\ref{show-reorganizing-dat}). The result in the first samples after this processing is shown in the table \ref{tab:new-features}.

@@

### Adding new variables
To finish the data preparation, new columns were added to the dataset (\ref{show-new-features-2}). Starting with the concrete class, for example if the compressive strength is between 25 and 30, it receives the class *C25*. The inclusion of the class was important because the compressive strength in *MPa* is a continuous variable, which will be used in the regression models, but the class as a discrete variable can provide another visualization of the data. The approximate mix of concrete was also added, which represents the proportions of aggregates (fine and coarse) for cement. Other proportions between the main ingredients were also added. The new variables are presented in the table ref{tab:new-features-table}.
```{r new-features-2, echo=F}
# Adding New features
concrete_class <- function(mpa){
  if (mpa >= 10) {
    s <- as.character(mpa)
    first <- substr(s, start = 1, stop = 1)
    second <- ifelse(substr(s, start = 2, stop = 2) >= 5, 5, 0)
  }
  else {
    first <- ""
    second <- "5"
  }
  paste("C", first, second, sep = "")
}
mix <- function(c, f_ag, c_ag){
  paste(
    1,
    round(f_ag/c, 0),
    round(c_ag/c, 0),
    sep = ":")
}
dat <- dat %>%
  mutate(class = sapply(day_28, concrete_class)) %>%
  mutate(class = as.factor(class)) %>%
  mutate(mix_app = factor(
    mix(cement, fine_aggregate, coarse_aggregate))) %>%
  mutate(`water_/_cement` = water / cement) %>%
  mutate(`fine_aggregate_/_cement` = fine_aggregate/cement) %>%
  mutate(`coarse_aggregate_/_cement` = coarse_aggregate/cement) %>%
  mutate(`fine_aggregate_/_coarse_aggregate` = fine_aggregate/coarse_aggregate) %>%
  mutate(`water_/_coarse_aggregate` = water/coarse_aggregate) %>%
  mutate(`water_/_fine_aggregate` = water/fine_aggregate)
lvl <- levels(dat$class)
dat$class <- factor(
  dat$class, 
  levels=c( "C5", sort(lvl[lvl!="C5"], decreasing=F)))
```
```{r new-features-table, echo=F}
# Table - New features
caption <- "New features"
colNames7 = c("ID", "Class","Approximated Mix",
             "Water / Cement", "Fine Ag. / Cement", 
             "Coarse Ag. / Cement", "Fine Ag. / Coarse Ag.",
             "Water / Coarse Ag.", "Water / Fine Ag.")
kable(
    head(dat[order(dat$id),][,c(1,15:22)]),
    col.names = colNames7,
    escape = F,
    booktabs = T,
    caption = caption,
    linesep = "\\addlinespace",
    align = "c",
    digits = 4
    ) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  column_spec(2, width = "1.5cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4:9, width = "1.7cm")
```

@@

### Descriptive statistics
The table \ref{tab:stat-summ} presents the statistical data of the continuous variables (\ ref {show-stat-summ}). The *Null* line represents the number of zeroed values for the ingredients, and the *NA* line represents the number of missing data. As the samples were filtered to maintain only sets of samples with known values of compressive strength at 28 days, the number of *NAs* is zero for that age. The figure \ref{fig:stat-summ-categorical} presents the statistical data of the discrete variables (\ref{show-stat-summ-categorical}).

@@

### Variables distribution
A figura \ref{fig:vars-distribution} mostra a distribuição das variáveis nas amostras (\ref{show-vars-distribution}). Foi calculado utilizando apenas os dados aos 28 dias. 
The figure\ref{fig:vars-distribution} shows the distribution of variables in the samples (\ref{show-vars-distribution}). It was calculated using data only at 28 days.

@@

### Principal component analysis
Na figura \ref{fig:pca}, utilizando uma classificação alternativa, foi realizada a análise de componente principal nos ingredientes das amostras (\ref{show-pca}). A classificação separa o concreto em 4 grupos de resistência a compressão diferentes, baixo até $20 MPa$, normal até $40 MPa$, médio até $70 MPa$ e alto acima disso. É possivel perceber que os grupos se sobrepõem, mas existem uma diferenciação entre o grupo alto e baixo.
In the figure v, using an alternative classification, the principal component analysis was performed on the ingredients (\ref{show-pca}). The classification separates concrete into 4 different compressive strength groups, low up to *20 MPa*, normal up to *40 MPa*, medium up to *70 MPa* and high above that. It is possible to notice that the groups overlap, but there is a differentiation between the high and low group.
```{r pca, echo=F, fig.height=3, fig.cap=cap}
# Figure - Principal component analysis on ingredients

@@

## Machine learning models
The development of machine learning models was carried out with the *caret* package [@caret] and based on @irizarry2019 e @Kuhn2008.

### Pre-processing and data separation
Como existe a variável categórica para o traço aproximado do concreto, foi realizada a conversão dessa variável em variáveis fictiícias (*dummy vars*) (\ref{show-dummy-var}), passando de 22 colunas (id, classe, resistência à compressão e mais 19 *features*) para 45 colunas, uma adição de 23 variáveis, uma para cada traço aproximado. 
As the approximate mix is a categorical variable, it was converted into dummy variables (\ref{show-dummy-var}), going from 22 columns (id, class, compressive strength and 19 more *features*) to 45 columns, an addition of 23 variables, one for each approximate mix.

@@

### Table First samples

>>>>

---
title:
  Concrete compressive strength prediction with machine learning
date: "April 22, 2020"
author: "Pedro Bernardino Alves Moreira"
indent: true
abstract: "Compressive strength is the main characteristic of concrete. The correct prediction of this parameter means cost and time reduction. This work built predictive models for 6 different ages of concrete samples (3, 7, 14, 28, 56, and 100 days). A set of data obtained in previous studies was used, a total of 1030 samples, with 9 variables: compressive strength, age, and 7 ingredients (water, cement, fine aggregate, coarse aggregate, fly ash, blast furnace slag, and superplasticizers). Another 6 variables were added to represent the proportions of the main ingredients in each sample (water/cement, fine aggregate/cement, coarse aggregate/cement, fine aggregate/coarse aggregate, water/coarse aggregate, and water/fine aggregate). The predictive models were developed in *R* language, using the *caret* package with the *Parallel Random Forest* algorithm and repeated cross-validation technique to optimize the parameters. The results were satisfactory and compatible with other studies using the same data set. The most important model, 28 days old, obtained *RMSE* of 4.717. The 3-day model obtained the best result, *RMSE* of 3.310. The worst result was the 56-day model, with *RMSE* of 5.939. The work showed that the compressive strength of concrete can be predicted. The choice of creating a model for each age, instead of using age as a predictor, allowed to get compatible results with the available data at each age. It was a promising alternative since good results were achieved by training with just one algorithm. This work facilitates exploration and new efforts to predict the compressive strength of concrete, it can be replicated using different algorithms or the combination of several."
output:
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc_depth: 2
bibliography: references.bib
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{supertabular}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
- \usepackage{caption}
- \usepackage{floatrow}
- \floatsetup[table]{capposition=top}
- \floatsetup[figure]{capposition=top}
- \captionsetup{options=chunk}
- \DeclareNewFloatType{chunk}{placement=H, fileext=chk, name=}
- \renewcommand{\thechunk}{\arabic{chunk}}
- \usepackage{indentfirst}
- \usepackage{sectsty} \sectionfont{\centering}
---
```{r setup, include=FALSE, cache=F}
# Install libraries, if not already installed
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(caret)) install.packages("caret")
if(!require(knitr)) install.packages("knitr")
if(!require(kableExtra)) install.packages("kableExtra")
if(!require(gdata)) install.packages("gdata")
if(!require(dplyr)) install.packages("dplyr")
if(!require(tidyr)) install.packages("tidyr")
if(!require(cowplot)) install.packages("cowplot")
if(!require(pastecs)) install.packages("pastecs")
if(!require(factoextra)) install.packages("factoextra")
if(!require(reshape2)) install.packages("reshape2")
if(!require(purrr)) install.packages("purrr")
if(!require(gridExtra)) install.packages("gridExtra")
if(!require(questionr)) install.packages("questionr")
# Load libraries
library(tidyverse)
library(caret)
library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(gdata)
library(reshape2)
library(cowplot)
library(pastecs)
library(factoextra)
library(purrr)
library(gridExtra)
library(questionr)
# Set global options
options(pillar.sigfig = 5)
options(knitr.kable.NA = '')
opts_knit$set(eval.after = "fig.cap")
opts_chunk$set(echo = TRUE, fig.align = 'center', cache=T, prompt=F, highlight=T )
oldSource <- knit_hooks$get("source")
knit_hooks$set(source = function(x, options) {
  x <- oldSource(x, options)
  x <- ifelse(!is.null(options$ref), paste0("\\label{", options$ref,"}", x), x)
  ifelse(!is.null(options$codecap), paste0("\\captionof{chunk}{", options$codecap,"}", x), x)
})
\newpage
\tableofcontents
\newpage
\listoffigures
\listoftables
\newpage

@@

## Obtaining the data
The data was downloaded from the University of California Irvine website [@downloadData] (\ref{show-download-data}). In total there are 1030 samples with 9 columns. The samples were renamed and an id column was added to facilitate data manipulation (\ref{show-rename-dat-cols}). The columns were reordered to put the new id column in the first position (\ref{show-reorder-dat}). The first samples can be viewed in the table \ref{tab:first-samples}.

@@

```{r download-data, echo=F, include=F}
# Data download
url_base <- "https://archive.ics.uci.edu"
url <-  "/ml/machine-learning-databases/concrete/compressive/Concrete_Data.xls"
download.file(paste0(url_base, url), "data.xls")
dat <- read.xls("data.xls")
n_inicial_samples <- nrow(dat)
colnames(dat)
```
```{r rename-dat-cols, echo=F}
# Renaming the columns
colnames(dat) <- c(
  "cement",
  "blast_furnace_slag",
  "fly_ash",
  "water",
  "superplasticizers",
  "coarse_aggregate",
  "fine_aggregate",
  "day",
  "mpa"
  )
dat$id <- seq.int(nrow(dat))
```
```{r reorder-dat, echo=F}
# Reordering the data
col_order <- c(
  "id",
  "cement",
  "blast_furnace_slag",
  "fly_ash",
  "water",
  "superplasticizers",
  "coarse_aggregate",
  "fine_aggregate",
  "day",
  "mpa"
)
dat <- dat[, col_order]
```
```{r col-names-and-units, echo=F}
# Defining column names and units
colNames <- c("ID", "Cement", "B.F.S.", "Fly ash", "Water",
             "Superp.", "C.Aggregate", "F.Aggregate", "Day", "Comp.Str.")
dfUnits <- c("", "$kg/m^3$", "$kg/m^3$","$kg/m^3$","$kg/m^3$",
             "$kg/m^3$","$kg/m^3$","$kg/m^3$","","$MPa$")
```
```{r first-samples, echo=F}
# Table - First samples
caption <- "First 6 samples"
kable(
    dat[1:6,],
    col.names = dfUnits,
    escape = F,
    booktabs = T,
    caption = caption,
    linesep = "\\addlinespace",
    align = "c"
    ) %>%
    add_header_above(header = colNames, line = F, align = "c") %>%
    kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

@@

```{r removing-duplicated-samples, echo=F}
# Removing duplicated samples
n_distinct_samples <- dat %>% select(-c(id)) %>% n_distinct()
n_duplicated_samples <- n_inicial_samples - n_distinct_samples
dat <- dat[!duplicated(select(dat, -c(id))),]
n_samples <- nrow(dat)
```
The data show the variables in the columns and samples in the rows. However it was found that some samples are identical in proportions of ingredients, changing only the value of age and compressive strength, for example, samples 653, 654, 678 and 681, shown in the table \ref{tab:similar-samples}.

@@

```{r similar-samples, echo=FALSE}
# Table - Samples with same composition
same_samples <- dat %>%
  filter(id %in% c(653, 678, 654, 681))
caption <- "Samples with same composition"
kable(
  same_samples[order(same_samples$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```
In addition, there are also samples with the same values and proportions of ingredients, but with different compressive strength, probably due to differences in the building process. This is the case, for example, of samples 472, 473 and 474, shown in the table  \ref{tab:similar-samples-2}.

@@

```{r similar-samples-2, echo=FALSE}
# Table - Same samples with different results
same_samples_2 <- dat %>%
  filter(id %in% c(472, 473, 474))
caption <- "Same samples with different results"
kable(
  same_samples_2[order(same_samples_2$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("scale_down"))
```
To facilitate the analysis of the samples, all samples that are the same in relation to the ingredients, have been assigned the same *id*. In addition, as the compressive strength at 28 days is the parameter to determine the grade of the concrete, only the elements containing that day among its samples were maintained. In the case of the same samples but with different results of compressive strength, the values were averaged. After all these changes (\ref{show-initial-data-cleaning}), the new total of samples was reduced to 970, containing 416 different settings for the proportions of ingredients.

@@

```{r similar-samples-same-id, echo=FALSE, include=TRUE}
# Table - Previous samples after processing
same_samples <- dat %>%
  filter(id == 653 | id == 472 & day == 28)
caption <- "Previous samples after processing"
kable(
  same_samples[order(same_samples$id, same_samples$day),],
  col.names = dfUnits,
  escape = F,
  booktabs = T,
  caption = caption,
  linesep = "\\addlinespace",
  align = "c",
  row.names = FALSE,
  digits = 2
  ) %>%
  add_header_above(header = colNames, line = F, align = "c") %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

### Age selection
As previously described, the main age for analysis of compressive strength is 28 days, but other ages can also be used to build predictive models. However, it is necessary to verify how relevant the data of these other ages are. Starting with the distribution of the samples in relation to each age  (\ref{show-boxplot}) shown in the figure \ref{fig:boxplot}.

@@

This hypothesis was verified using the principal component analysis method, applied to samples of these 3 ages (\ref{show-pca-90-91-100}). The figure \ref{fig:pca-90-91-100} shows how the samples relate to each other (which are similar or different) and revealed how each variable contributes to the analysis. The first two dimensions represent 37% and 24% $ respectively of the variance.

@@

```{r pca-90-91-100, echo=F, include=T, fig.cap = cap, fig.height=3.5}
# Figure - Principal component analysis - 90, 91 e 100 days
dat_90_91_100 <- dat %>%
  ungroup() %>%
  filter(day %in% c(90, 91, 100)) %>%
  select(-c(id, mpa))
cap <- "Principal component analysis - 90, 91 e 100 days"
colnames(dat_90_91_100) <- c(
   "Cem.", "B.F.S.", "Fly.A.","Water","Sup.","C.Ag.","F.Ag.","day")
pca <- prcomp(select(dat_90_91_100, -c(day)), scale = TRUE)
habillage <- dat_90_91_100$day
fviz_pca_biplot(
  pca,
  geom.ind = "point",
  habillage=habillage,
  addEllipses = TRUE,
  ellipse.level=0.75) +
  ggtitle("") +
  theme_bw() +
  coord_cartesian(xlim = c(-3, 3.5), ylim = c(3, -5))
```
Another important point considered, of the concrete nature itself, is the fact that the growth rate of its compressive strength decreases with time, reaching a certain stability value. The figure \ref{fig:mpa-on-time} shows the compressive strength over the days for samples with more than 5 data, that is, data available for at least 6 different ages (\ref{show-mpa-on-time}).

@@

```{r mpa-on-time, echo=F, include=T, fig.cap=cap, fig.height=3}
# Figure - Compressive strength through time
cap <- "Compressive strength through time"
ylabel <- "MPa"
xlabel <- "Age (days)"
dat_duplicated_only <- dat %>%
  group_by(id) %>%
  filter(n()>5) %>%
  select(id, mpa, day)
dat_duplicated_only %>%
  ggplot(aes(day, mpa, fill=id, alpha = 0.5)) +
  geom_line() +
  xlab(xlabel) +
  ylab(ylabel) +
  theme_bw() +
  theme(legend.position = "none")
```
For the reasons presented in the figures \ref{fig:boxplot}, \ref{fig:pca-90-91-100} and \ref{fig:mpa-on-time}, it was considered that the ages of 90, 91 and 100 days can be grouped to improve reading and decrease sample noise. They were converted to the same value, the age of 100 days was chosen (\ref{show-join-90-91-100}). As shown in the figure \ref{fig:mpa-on-time}, the resistance only increases, after 100 days the resistance to compression will be greater than or equal to the value of 90 or 91 days.

@@

```{r join-90-91-100, echo=F}
# Joining 90, 91 and 100 days data
ind_90 <- dat$id[which(dat$day == "90")]
ind_91 <- dat$id[which(dat$day == "91")]
ind_100 <- dat$id[which(dat$day == "100")]
sum_duplicated <- sum(duplicated(c(ind_90, ind_91, ind_100))) # 0
dat <- dat %>%
  ungroup() %>%
  mutate(day = ifelse(day %in% c(91, 90), 100, day))
```
Another topic analyzed in the selection of ages was the observed frequency of each age value after this transformation from 90, 91 in 100 days, shown in the figure \ref{fig:freq-ages}. Some values of days have very low concentrations of samples, at the risk of damaging more than helping to create the models, so they were removed (\ref{show-remove-ages-lower-50}). The criterion adopted was to maintain only ages with a frequency greater than 50, only the values of 3, 7, 14, 28, 56 and 100 days.

@@

### Data reorganization
The samples were grouped to maintain only one sample from each set of configuration of the proportions of the ingredients, adding new variables/columns for resistance at each age (\ref{show-reorganizing-dat}). The result in the first samples after this processing is shown in the table \ref{tab:new-features}.

@@

### Adding new variables
To finish the data preparation, new columns were added to the dataset (\ref{show-new-features-2}). Starting with the concrete class, for example if the compressive strength is between 25 and 30, it receives the class *C25*. The inclusion of the class was important because the compressive strength in *MPa* is a continuous variable, which will be used in the regression models, but the class as a discrete variable can provide another visualization of the data. The approximate mix of concrete was also added, which represents the proportions of aggregates (fine and coarse) for cement. Other proportions between the main ingredients were also added. The new variables are presented in the table \ref{tab:new-features-table}.
```{r new-features-2, echo=F}
# Adding New features
concrete_class <- function(mpa){
  if (mpa >= 10) {
    s <- as.character(mpa)
    first <- substr(s, start = 1, stop = 1)
    second <- ifelse(substr(s, start = 2, stop = 2) >= 5, 5, 0)
  }
  else {
    first <- ""
    second <- "5"
  }
  paste("C", first, second, sep = "")
}
mix <- function(c, f_ag, c_ag){
  paste(
    1,
    round(f_ag/c, 0),
    round(c_ag/c, 0),
    sep = ":")
}
dat <- dat %>%
  mutate(class = sapply(day_28, concrete_class)) %>%
  mutate(class = as.factor(class)) %>%
  mutate(mix_app = factor(
    mix(cement, fine_aggregate, coarse_aggregate))) %>%
  mutate(`water_/_cement` = water / cement) %>%
  mutate(`fine_aggregate_/_cement` = fine_aggregate/cement) %>%
  mutate(`coarse_aggregate_/_cement` = coarse_aggregate/cement) %>%
  mutate(`fine_aggregate_/_coarse_aggregate` = fine_aggregate/coarse_aggregate) %>%
  mutate(`water_/_coarse_aggregate` = water/coarse_aggregate) %>%
  mutate(`water_/_fine_aggregate` = water/fine_aggregate)
lvl <- levels(dat$class)
dat$class <- factor(
  dat$class, 
  levels=c( "C5", sort(lvl[lvl!="C5"], decreasing=F)))
```
```{r new-features-table, echo=F}
# Table - New features
caption <- "New features"
colNames7 = c("ID", "Class","Approximated Mix",
             "Water / Cement", "Fine Ag. / Cement", 
             "Coarse Ag. / Cement", "Fine Ag. / Coarse Ag.",
             "Water / Coarse Ag.", "Water / Fine Ag.")
kable(
    head(dat[order(dat$id),][,c(1,15:22)]),
    col.names = colNames7,
    escape = F,
    booktabs = T,
    caption = caption,
    linesep = "\\addlinespace",
    align = "c",
    digits = 4
    ) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>%
  column_spec(2, width = "1.5cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4:9, width = "1.7cm")
```

@@

### Descriptive statistics
The table \ref{tab:stat-summ} presents the statistical data of the continuous variables (\ref {show-stat-summ}). The *Null* line represents the number of zeroed values for the ingredients, and the *NA* line represents the number of missing data. As the samples were filtered to maintain only sets of samples with known values of compressive strength at 28 days, the number of *NAs* is zero for that age. The figure \ref{fig:stat-summ-categorical} presents the statistical data of the discrete variables (\ref{show-stat-summ-categorical}).

@@

### Variables distribution
The figure\ref{fig:vars-distribution} shows the distribution of variables in the samples (\ref{show-vars-distribution}). It was calculated using data only at 28 days.

@@

### Principal component analysis
In the figure \ref{fig:pca}, using an alternative classification, the principal component analysis was performed on the ingredients (\ref{show-pca}). The classification separates concrete into 4 different compressive strength groups, low up to *20 MPa*, normal up to *40 MPa*, medium up to *70 MPa* and high above that. It is possible to notice that the groups overlap, but there is a differentiation between the high and low group.
```{r pca, echo=F, fig.height=3, fig.cap=cap}
# Figure - Principal component analysis on ingredients

@@

## Machine learning models
The development of machine learning models was carried out with the *caret* package [@caret] and based on @irizarry2019 and @Kuhn2008.

### Pre-processing and data separation
As the approximate mix is a categorical variable, it was converted into dummy variables (\ref{show-dummy-var}), going from 22 columns (id, class, compressive strength and 19 more *features*) to 45 columns, an addition of 23 variables, one for each approximate mix.

@@

### Table - First samples
